---
title: "OOPS_TMT_Statistical_testing"
output: html_notebook
---
30-03-2023
Notebook for statistical testing of my OOPS TMT experiment with toxic ASO vs non-toxic ASO vs vehicle

```{r, message=FALSE}
library(camprotR)
library(ggplot2)
library(MSnbase)
library(DEqMS)
library(limma)
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(biobroom)
library(uniprotREST)
library(here)

```

First, we combine the Input and RBP into one MSnset
```{r}


# read in the MsnSets with input and OOPS quantification at protein-level
exp1 <- readRDS(here('Results/input_tmt_protein.rds'))

exp2 <- readRDS(here('Results/OOPS_tmt_protein.rds'))

# identify the common proteins
common_proteins <- intersect(rownames(exp1), rownames(exp2))

# define a new experimental condition to distinguish input vs OOPS
pData(exp1)$material='Input'
pData(exp2)$material='RBP'

# Reduce the feature data to remove columns we don't need and that are specific
# to one TMT plex or the other
fData(exp1) <- fData(exp1)[,c('Master.Protein.Accessions', "Master.Protein.Descriptions"),drop=FALSE]
fData(exp2) <- fData(exp2)[,c('Master.Protein.Accessions', "Master.Protein.Descriptions"),drop=FALSE]

expInputRBP <- MSnbase::combine(exp1[common_proteins,], exp2[common_proteins,])
```

Starting with modeling Toxic vs non-toxic ASO
```{r}
# Filtering to remove the samples which are Vehicle treatment
data_to_model <- expInputRBP[,pData(expInputRBP)$Treatment!='Vehicle'] 

#assign input and RBPs as factors in the material column of the data_to_model dataset.
material <- factor(data_to_model$material, levels=c('Input', 'RBP'))

toxic <- factor(data_to_model$Treatment, levels=c('Safe', 'Toxic'))


design <- model.matrix(~material*toxic)

# And then usual lines to perform linear modeling with limma...
```

Build the run_limma function
```{r}

run_limma <- function(data_to_model,
                      design){

  # run limma
  lmfit <- data_to_model %>%
    filterNA() %>%
    lmFit(design) %>%
    eBayes(trend=TRUE)

  # plot model vs mean variance
  plotSA(lmfit, main="Final model: mean-variance trend")

  return(lmfit)
}

summariseLmFit <- function(lmfit, coefficient){
  # obtain results table
  lmResults <- topTable(lmfit, coef=coefficient, number=Inf)



  # add max_diff column if required
  if(is.integer(coefficient) & length(coefficient)>1){
    logfc_col='max_diff'
    message('Multiple numerical coffficients. Adding a "max_diff" column')
    lmResults['max_diff'] <- lmResults %>%
      select(all_of(gsub(':', '.', colnames(coefficients(lmfit))[coefficient]))) %>%
      apply(MARGIN=1, function(x) x[which(abs(x)==max(abs(x)))])
  } else logfc_col='logFC'

  message('At 1% FDR')
  print(table(ifelse(lmResults$adj.P.Val<0.01, 'Significant', 'Not significant'),
              ifelse(lmResults[[logfc_col]]>0, 'Increase', 'Decrease')))

  message('At 20% FDR')
  print(table(ifelse(lmResults$adj.P.Val<0.2, 'Significant', 'Not significant'),
              ifelse(lmResults[[logfc_col]]>0, 'Increase', 'Decrease')))

  return(lmResults)
}
```

Build the volcano_plot function

```{r}


volcano_plot <- function(lmResultsObj,
                         p_col='P.Value',
                         adj_p_col='adj.P.Val',
                         adj.P.Val_threshold=0.05,
                         logFC_threshold=2,
                         xcol='logFC',
                         xlabel){
  lmResultsObj %>%
    mutate(sig=!!sym(adj_p_col)<adj.P.Val_threshold,
           large_effect=abs(!!sym(xcol))>logFC_threshold) %>%
    mutate(label=ifelse(sig & large_effect, gsub('(.*GN=| PE=.*)', '', Master.Protein.Descriptions), NA)) %>%
    ggplot(aes(!!sym(xcol), -log10(!!sym(p_col)), colour=sig)) +
    geom_point(aes(size=sig)) +
    theme_camprot(base_size=15, border=FALSE) +
    scale_colour_manual(values=c('grey', get_cat_palette(1)), guide=FALSE) +
    scale_size_manual(values=c(0.5, 2), guide=FALSE) +
    xlab(xlabel) +
    ylab('-log10(p-value)') +
    theme(strip.background=element_blank()) +
    ggrepel::geom_label_repel(aes(label=label), show.legend=FALSE)
}
```

```{r}
limma_fit <- run_limma(data_to_model, design)
lmResults <- summariseLmFit(limma_fit, coefficient='materialRBP:toxicToxic')

volcano <- volcano_plot(lmResults,
                          adj.P.Val_threshold=0.1,
                          logFC_threshold=0.9,
                          xlabel='Effect of Toxic ASO on RNA binding (log2)')

#define function build_results_list
build_results_list <- function(data_to_model, design, limma_fit, results, volcano){
  list('data'=data_to_model,
       'design'=design,
       'limma_fit'=limma_fit,
       'results'=results,
       'volcano'=volcano)
}

lm_results_toxic_vs_non_toxic <- build_results_list(data_to_model,
                                 design,
                                 limma_fit,
                                 lmResults,
                                 volcano)

print(lm_results_toxic_vs_non_toxic$volcano)


```



```{r}


# lmResults is ordered by p-value, so first protein has smallest p-value.
poi <- lmResults$Master.Protein.Accessions[1]
print(poi)

data_long_poi <- tidy(data_to_model, addPheno=TRUE) %>%  # get data in long form
  filter(protein==poi) # and filter to one protein of interest

 # to extract quantification and experimental information in tidy format
p <- ggplot(data_long_poi,
            aes(x=Treatment, y=value)) + 
  geom_point(size=3) +
  # add the mean
  stat_summary(fun = mean, size=3,
               geom = "point",
               colour=get_cat_palette(2)[2]) +
  # add the 95% CI
  stat_summary(fun.data = mean_cl_normal,
               geom = "errorbar", width=0.2,
               colour=get_cat_palette(2)[2],
              fun.args = list(mult = 1.96)) + # defaults to mean +/- 1 SE. Updated to ~95% CI
  facet_wrap(~material, scales='free_y') + # Allow different y-axis ranges for facets (Input & RBP)
  theme_bw(base_size=15) +
  theme(strip.background=element_blank()) +
  ylab('Abundance (log2)')

print(p)
```
OK, so the protein with the smallest p-value does at least look like it might have a change in RNA binding when inspected individually. Clearly though, the difference is too small to be significant when we are testing all proteins together.


------------------------------------------
limma didn't look particularly interesting so Tom has suggested using DEqMS at the PSM level instead.
First I need to import the PSM files for both input and for OOPS samples.
Then I need to combine them with the limma adjusted objects.



Import PSM files
```{r}

#OOPS samples

OOPS_tmt_psm_res <- readRDS(here('Results/OOPS_psm_filt.rds'))

#Inputs

Input_tmt_psm_res <- readRDS(here('Results/input_psm_filt.rds'))



```


Import protein files in MSnSet format
```{r}

OOPS_tmt_prot_res <- readRDS(here('Results/OOPS_tmt_protein.rds'))

Input_tmt_prot_res <- readRDS(here('Results/input_tmt_protein.rds'))
```

Now tidy them with phenotype data and filter out the vehicle treated-samples
```{r}
OOPS_protein_tidy <- OOPS_tmt_prot_res %>%
  biobroom::tidy.MSnSet(addPheno=TRUE) %>% # addPheno=TRUE adds the pData so we have the sample information too
  filter(is.finite(value))

OOPS_protein_tidy <- filter(OOPS_protein_tidy, Treatment!='Vehicle')

Input_protein_tidy <- Input_tmt_prot_res %>%
  biobroom::tidy.MSnSet(addPheno=TRUE) %>% # addPheno=TRUE adds the pData so we have the sample information too
  filter(is.finite(value))

Input_protein_tidy <- filter(Input_protein_tidy, Treatment!='Vehicle')
```

now get the minimum psm count for input and OOPS
```{r}
# Obtain the min peptide count across the samples and determine the minimum value across
# samples
#first OOPS
OOPS_min_psm_count <- camprotR::count_features_per_protein(OOPS_tmt_psm_res) %>%
  merge(OOPS_protein_tidy,
        by.x=c('Master.Protein.Accessions', 'sample'),
        by.y=c('protein', 'sample')) %>%
  group_by(Master.Protein.Accessions) %>%
  summarise(min_psm_count = min(n))

#######Now input
input_min_psm_count <- camprotR::count_features_per_protein(Input_tmt_psm_res) %>%
  merge(Input_protein_tidy,
        by.x=c('Master.Protein.Accessions', 'sample'),
        by.y=c('protein', 'sample')) %>%
  group_by(Master.Protein.Accessions) %>%
  summarise(min_psm_count = min(n))


#combine input and OOPS psm_counts 

min_psm_count_input_rbp <- bind_rows(input_min_psm_count, OOPS_min_psm_count) %>% group_by(Master.Protein.Accessions) %>% summarise(min_psm_count=min(min_psm_count))

```


Now that we have PSM counts for input and OOPS, add to limma object
```{r}

#checking number of rows for between objects 
nrow(data_to_model)
nrow(min_psm_count_input_rbp)

#showing differences between the two objects 
setdiff(rownames(data_to_model),
        min_psm_count_input_rbp$Master.Protein.Accessions)

psm_counts <-input_min_psm_count %>% tibble::remove_rownames() %>%
  tibble::column_to_rownames('Master.Protein.Accessions') 

limma_fit$count = psm_counts[common_proteins,'min_psm_count']
```



```{r}

deqms_fit = spectraCounteBayes(limma_fit)
# Diagnostic plots
VarianceBoxplot(deqms_fit,n=30,main="TMT15plex dataset",xlab="PSM count")
VarianceScatterplot(deqms_fit,,main="TMT15plex dataset",xlab="PSM count")

```

```{r}
deqms_results <- outputResult(deqms_fit, coef_col=4)

lm_results_toxic_vs_non_toxic$deqms_fit <- deqms_fit_safe
lm_results_toxic_vs_non_toxic$deqms_results <- deqms_results_safe

table(deqms_results$sca.adj.pval<0.01)


deqms_results %>%
  ggplot(aes(x = logFC, y = -log10(sca.P.Value), colour = sca.adj.pval < 0.01)) +
  geom_point(size=0.5) +
  theme_camprot(border=FALSE, base_size=15) +
  scale_colour_manual(values = c('grey', get_cat_palette(2)[2]), name = '2x vs 1x Sig.') +
  labs(x = 'Toxic vs Non-Toxic (Log2)', y = '-log10(p-value)') 


p <- volcano_plot(deqms_results, xlabel='Toxic vs Non-Toxic (Log2)',
             p_col='P.Value', adj_p_col='adj.P.Val') + ggtitle('limma')
print(p)

p <- volcano_plot(deqms_results, xlabel='Toxic vs Non-Toxic (Log2)',
             p_col='sca.P.Value', adj_p_col='sca.adj.pval') + ggtitle('DEqMS')
print(p)
```




---------------------------------------------------
Let's try looking at any differences between my L2K and safe ASO

```{r}
#L2k vs Safe ASO statistical testing


# Remove the samples which are Toxic ASO treated
data_to_model2 <- expInputRBP[,pData(expInputRBP)$Treatment!='Toxic'] 

#assign input and RBPs as factors in the material column of the data_to_model dataset.
material2 <- factor(data_to_model2$material, levels=c('Input', 'RBP'))

safe <- factor(data_to_model2$Treatment, levels=c('Safe', 'Vehicle'))

design2 <- model.matrix(~material2*safe)

# And then usual lines to perform linear modeling with limma...
```



```{r}
limma_fit_safe <- run_limma(data_to_model2, design2)
lmResults_safe <- summariseLmFit(limma_fit_safe, coefficient='material2RBP:safeVehicle')

volcano_safe <- volcano_plot(lmResults_safe,
                          adj.P.Val_threshold=0.1,
                          logFC_threshold=0.9,
                          xlabel='Effect of Safe ASO on RNA binding (log2)')


lm_results_aso_vs_vehicle <- build_results_list(data_to_model2,
                                 design2,
                                 limma_fit_safe,
                                 lmResults_safe,
                                 volcano_safe)

print(lm_results_aso_vs_vehicle$volcano)


```

```{r}


# lmResults is ordered by p-value, so first protein has smallest p-value.
poi_safe <- lmResults_safe$Master.Protein.Accessions[1]
print(poi_safe)

data_long_poi_safe <- tidy(data_to_model2, addPheno=TRUE) %>%  # get data in long form
  filter(protein==poi_safe) # and filter to one protein of interest

 # to extract quantification and experimental information in tidy format
p <- ggplot(data_long_poi_safe,
            aes(x=Treatment, y=value)) + 
  geom_point(size=3) +
  # add the mean
  stat_summary(fun = mean, size=3,
               geom = "point",
               colour=get_cat_palette(2)[2]) +
  # add the 95% CI
  stat_summary(fun.data = mean_cl_normal,
               geom = "errorbar", width=0.2,
               colour=get_cat_palette(2)[2],
              fun.args = list(mult = 1.96)) + # defaults to mean +/- 1 SE. Updated to ~95% CI
  facet_wrap(~material, scales='free_y') + # Allow different y-axis ranges for facets (Input & RBP)
  theme_bw(base_size=15) +
  theme(strip.background=element_blank()) +
  ylab('Abundance (log2)')

print(p)
```
Now that we have PSM counts for input and OOPS, add to limma object
```{r}

limma_fit_safe$count = psm_counts[common_proteins,'min_psm_count']

deqms_fit_safe = spectraCounteBayes(limma_fit_safe)
# Diagnostic plots
VarianceBoxplot(deqms_fit_safe,n=30,main="TMT15plex dataset",xlab="PSM count")
VarianceScatterplot(deqms_fit_safe,main="TMT15plex dataset",xlab="PSM count")

```

```{r}
deqms_results_safe <- outputResult(deqms_fit_safe, coef_col=4)

lm_results_aso_vs_vehicle$deqms_fit <- deqms_fit_safe
lm_results_aso_vs_vehicle$deqms_results <- deqms_results_safe


table(deqms_results_safe$sca.adj.pval<0.01)

deqms_results_safe %>%
  ggplot(aes(x = logFC, y = -log10(sca.P.Value), colour = sca.adj.pval < 0.01)) +
  geom_point(size=0.5) +
  theme_camprot(border=FALSE, base_size=15) +
  scale_colour_manual(values = c('grey', get_cat_palette(2)[2]), name = '2x vs 1x Sig.') +
  labs(x = 'Toxic vs Non-Toxic (Log2)', y = '-log10(p-value)') 

p <- volcano_plot(deqms_results_safe, xlabel='Toxic vs Non-Toxic (Log2)',
             p_col='P.Value', adj_p_col='adj.P.Val') + ggtitle('limma')
print(p)

p <- volcano_plot(deqms_results_safe, xlabel='Toxic vs Non-Toxic (Log2)',
             p_col='sca.P.Value', adj_p_col='sca.adj.pval') + ggtitle('DEqMS')
print(p)
```
Save out objects for the two tests for use in downstream notebooks.
```{r}
saveRDS(lm_results_toxic_vs_non_toxic, here('Results/lm_results_toxic_vs_non_toxic.rds'))
saveRDS(lm_results_aso_vs_vehicle, here('Results/lm_results_aso_vs_vehicle.rds'))
```


