---
title: "OOPMS_TMT_Statistical_Testing"
output: html_notebook
---
---
title: "OOPS_TMT_Statistical_Testing"
output: html_notebook
---
30-03-2023
Notebook for statistical testing of my OOPS TMT experiment with toxic ASO vs non-toxic ASO vs vehicle

```{r, message=FALSE}
library(camprotR)
library(ggplot2)
library(MSnbase)
library(DEqMS)
library(limma)
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(biobroom)
library(uniprotREST)
library(here)

```

```{r}


# prot_res_irs_norm is created in the previous vignette and is part of the snd1.kd.upr.rbp package
exp1 <- readRDS(here('Results/input_tmt_protein.rds'))

exp2 <- readRDS(here('Results/OOPS_tmt_protein.rds'))

common_proteins <- intersect(rownames(exp1), rownames(exp2))

#Combine the Input and RBP into one MSnset
pData(exp1)$material='Input'
pData(exp2)$material='RBP'

fData(exp1) <- fData(exp1)[,c('Master.Protein.Accessions', "Master.Protein.Descriptions"),drop=FALSE]

fData(exp2) <- fData(exp2)[,c('Master.Protein.Accessions', "Master.Protein.Descriptions"),drop=FALSE]

expInputRBP <- MSnbase::combine(exp1[common_proteins,], exp2[common_proteins,])
```

```{r}
data_to_model <- expInputRBP[,pData(expInputRBP)$Treatment!='Vehicle'] # or some equivalent filtering to remove the samples which are Vehicle treatment

#assign input and RBPs as factors in the material column of the data_to_model dataset.
material <- factor(data_to_model$material, levels=c('Input', 'RBP'))

toxic <- factor(data_to_model$Treatment, levels=c('Safe', 'Toxic'))


design <- model.matrix(~material*toxic)

# And then usual lines to perform linear modeling with limma...
```

Build the run_limma function
```{r}

run_limma <- function(data_to_model,
                      design,
                      coefficient){

  # run limma
  lmfit <- data_to_model %>%
    filterNA() %>%
    lmFit(design) %>%
    eBayes(trend=TRUE)

  # plot model vs mean variance
  plotSA(lmfit, main="Final model: mean-variance trend")

  # obtain results table
  lmResults <- topTable(lmfit, coef=coefficient, number=Inf)



  # add max_diff column if required
  if(is.integer(coefficient) & length(coefficient)>1){
    logfc_col='max_diff'
    message('Multiple numerical coffficients. Adding a "max_diff" column')
    lmResults['max_diff'] <- lmResults %>%
      select(all_of(gsub(':', '.', colnames(coefficients(lmfit))[coefficient]))) %>%
      apply(MARGIN=1, function(x) x[which(abs(x)==max(abs(x)))])
  } else logfc_col='logFC'

  message('At 1% FDR')
  print(table(ifelse(lmResults$adj.P.Val<0.01, 'Significant', 'Not significant'),
              ifelse(lmResults[[logfc_col]]>0, 'Increase', 'Decrease')))

  message('At 20% FDR')
  print(table(ifelse(lmResults$adj.P.Val<0.2, 'Significant', 'Not significant'),
              ifelse(lmResults[[logfc_col]]>0, 'Increase', 'Decrease')))

  return(lmResults)
}
```

build the volcano_plot function

```{r}
volcano_plot <- function(lmResultsObj,
                         adj.P.Val_threshold=0.05,
                         logFC_threshold=2,
                         xcol='logFC',
                         xlabel){
  lmResultsObj %>%
    mutate(sig=adj.P.Val<adj.P.Val_threshold,
           large_effect=abs(!!sym(xcol))>logFC_threshold) %>%
    mutate(label=ifelse(sig & large_effect, gsub('(.*GN=| PE=.*)', '', Master.Protein.Descriptions), NA)) %>%
    ggplot(aes(!!sym(xcol), -log10(P.Value), colour=sig)) +
    geom_point(aes(size=sig)) +
    theme_camprot(base_size=15, border=FALSE) +
    scale_colour_manual(values=c('grey', get_cat_palette(1)), guide=FALSE) +
    scale_size_manual(values=c(0.5, 2), guide=FALSE) +
    xlab(xlabel) +
    ylab('-log10(p-value)') +
    theme(strip.background=element_blank()) +
    ggrepel::geom_label_repel(aes(label=label), show.legend=FALSE)
}
```


Plot the limma volcano

```{r}

lmResults <- run_limma(data_to_model, design, coefficient='materialRBP:toxicToxic')

volcano <- volcano_plot(lmResults,
                          adj.P.Val_threshold=0.1,
                          logFC_threshold=0.9,
                          xlabel='Effect of Toxic ASO on RNA binding (log2)')

#define function build_results_list
build_results_list <- function(data_to_model, design, results, volcano){
  list('data'=data_to_model,
       'design'=design,
       'results'=results,
       'volcano'=volcano)
}

lm_results <- build_results_list(data_to_model,
                                   design,
                                   lmResults,
                                   volcano)
print(lm_results$volcano)


```

Limma showed no significant RBPs

plot data to model, looking at a specific protein to see if there is any variability between the samples.

```{r}
library(biobroom)
p<- tidy(data_to_model, addPheno=TRUE)

q<- p %>% filter(protein=="Q15050")

q$value <- log2(q$value)


 # to extract quantification and experimental information in tidy format
s<-ggplot(q, aes(x=factor(sample, level=c('127N','128C', '130N', '131C',  '133N','130C','127C','129N','132N','133C','Input127N','Input128C', 'Input130N', 'Input131C','Input133N','Input130C','Input127C','Input129N','Input132N', 'Input133C')), y=value)) + 
  geom_dotplot(binaxis='y', stackdir='center', aes(fill=Treatment))

s
```

------------------------------------------
limma didn't look particularly interesting so Tom has suggested using DEqMS at the PSM level instead.
First i need to import the PSM files for both input and for OOPS samples.
Then i need to combine them with the limma adjusted objects.



Import PSM files
```{r}

#OOPS samples

OOPS_tmt_psm_res <- readRDS(here('Results/OOPS_psm_filt.rds'))

#Inputs

Input_tmt_psm_res <- readRDS(here('Results/input_psm_filt.rds'))



```


Import protein files in tidy format
```{r}

OOPS_tmt_prot_res <- readRDS(here('Results/OOPS_tmt_protein.rds'))

Input_tmt_prot_res <- readRDS(here('Results/input_tmt_protein.rds'))
```

Now tidy them with phenotype data and filter out the vehicle treated-samples
```{r}
OOPS_protein_tidy <- OOPS_tmt_prot_res %>%
  biobroom::tidy.MSnSet(addPheno=TRUE) %>% # addPheno=TRUE adds the pData so we have the sample information too
  filter(is.finite(value))

OOPS_protein_tidy <- filter(OOPS_protein_tidy, Treatment!='Vehicle')

#Inputs

Input_protein_tidy <- Input_tmt_prot_res %>%
  biobroom::tidy.MSnSet(addPheno=TRUE) %>% # addPheno=TRUE adds the pData so we have the sample information too
  filter(is.finite(value))

Input_protein_tidy <- filter(Input_protein_tidy, Treatment!='Vehicle')


```

now get the minimum psm count for input and OOPS
```{r}
# Obtain the min peptide count across the samples and determine the minimum value across
# samples
#first OOPS
OOPS_min_psm_count <- camprotR::count_features_per_protein(OOPS_tmt_psm_res) %>%
  merge(OOPS_protein_tidy,
        by.x=c('Master.Protein.Accessions', 'sample'),
        by.y=c('protein', 'sample')) %>%
  group_by(Master.Protein.Accessions) %>%
  summarise(min_psm_count = min(n))

#######Now input
input_min_psm_count <- camprotR::count_features_per_protein(Input_tmt_psm_res) %>%
  merge(Input_protein_tidy,
        by.x=c('Master.Protein.Accessions', 'sample'),
        by.y=c('protein', 'sample')) %>%
  group_by(Master.Protein.Accessions) %>%
  summarise(min_psm_count = min(n))


#combine input and OOPS psm_counts 

min_psm_count_input_rbp <- bind_rows(input_min_psm_count, OOPS_min_psm_count) %>% group_by(Master.Protein.Accessions) %>% summarise(min_psm_count=min(min_psm_count))

limma_fit$count <- min_psm_count_input_rbp$min_psm_count

```


Now that we have PSM counts for input and OOPS, add to limma object
```{r}

#checking number of rows for between objects 
nrow(data_to_model)
nrow(min_psm_count_input_rbp)

#showing differences between the two objects 
setdiff(rownames(data_to_model),
        min_psm_count_input_rbp$Master.Protein.Accessions)


```



```{r}
#take common proteins between input and OOPS protein Rds and input min psm count, use this object for deqms
fit1 <- lmFit(data_to_model, design)
#fit2 <- contrasts.fit(fit1,contrasts = contrast)
fit3 <- eBayes(fit1, trend = TRUE)


fit3$count = tibble::column_to_rownames(
   input_min_psm_count,'Master.Protein.Accessions')[common_proteins,'min_psm_count']

fit4 = spectraCounteBayes(fit3)
VarianceBoxplot(fit4,n=30,main="TMT15plex dataset",xlab="PSM count")


efit_deqms <- suppressWarnings(spectraCounteBayes(fit4))


# Diagnostic plots
VarianceBoxplot(efit_deqms, n = 30, xlab = "PSMs")

```

```{r}
deqms_results <- outputResult(efit_deqms, coef_col=4)


table(deqms_results$sca.adj.pval<0.01)
#> 
#> FALSE  TRUE 
#>  33    863


deqms_results %>%
  ggplot(aes(x = logFC, y = -log10(sca.P.Value), colour = sca.adj.pval < 0.01)) +
  geom_point(size=0.5) +
  theme_camprot(border=FALSE, base_size=15) +
  scale_colour_manual(values = c('grey', get_cat_palette(2)[2]), name = '2x vs 1x Sig.') +
  labs(x = 'Toxic vs Non-Toxic (Log2)', y = '-log10(p-value)') 



```
```{r}

```


Testing Errors suggested by Tom
```{r}
table(lmResults$count)

table(is.na(lmResults$count))

table(is.na(exprs(data_to_model)))

table(is.na(input_min_psm_count))

rowSums(is.na(exprs(data_to_model)))

rowSums(is.na(input_min_psm_count))
```

```{r}
volcano_plot <- function(lmResults,
                         adj.P.Val_threshold=0.05,
                         logFC_threshold=2,
                         xcol='logFC',
                         xlabel){
  lmResults %>%
    mutate(sig=adj.P.Val<adj.P.Val_threshold,
           large_effect=abs(!!sym(xcol))>logFC_threshold) %>%
    mutate(label=ifelse(sig & large_effect, gsub('(.*GN=| PE=.*)', '', Master.Protein.Descriptions), NA)) %>%
    ggplot(aes(!!sym(xcol), -log10(P.Value), colour=sig)) +
    geom_point(aes(size=sig)) +
    theme_camprot(base_size=15, border=FALSE) +
    scale_colour_manual(values=c('grey', get_cat_palette(1)), guide=FALSE) +
    scale_size_manual(values=c(0.5, 2), guide=FALSE) +
    xlab(xlabel) +
    ylab('-log10(p-value)') +
    theme(strip.background=element_blank()) +
    ggrepel::geom_label_repel(aes(label=label), show.legend=FALSE)
}


volcano_plot <- function(lmResults,
                         adj.P.Val_threshold=0.05,
                         logFC_threshold=2,
                         xcol='logFC',
                         xlabel){
  lmResults %>%
    mutate(sig=adj.P.Val<adj.P.Val_threshold,
           large_effect=abs(!!sym(xcol))>logFC_threshold) %>%
    mutate(label=ifelse(sig & large_effect, gsub('(.*GN=| PE=.*)', '', Master.Protein.Descriptions), NA)) %>%
    ggplot(aes(!!sym(xcol), -log10(P.Value), colour=sig)) +
    geom_point(aes(size=sig)) +
    theme_camprot(base_size=15, border=FALSE) +
    scale_colour_manual(values=c('grey', get_cat_palette(1)), guide=FALSE) +
    scale_size_manual(values=c(0.5, 2), guide=FALSE) +
    xlab(xlabel) +
    ylab('-log10(p-value)') +
    theme(strip.background=element_blank()) +
    ggrepel::geom_label_repel(aes(label=label), show.legend=FALSE)
}



volcano <- volcano_plot(lmResults,
                          adj.P.Val_threshold=0.1,
                          logFC_threshold=0.9,
                          xlabel='Effect of Toxic ASO on RNA binding (log2)')

#define function build_results_list
build_results_list <- function(lmResults, design, results, volcano){
  list('data'=lmResults,
       'design'=design,
       'results'=results,
       'volcano'=volcano)
}

lmResults <- build_results_list(lmResults,
                                   design,
                                   lmResults,
                                   volcano)
print(lmResults$volcano)
```

---------------------------------------------------
Let's try looking at any differences between my L2K and safe ASO

```{r}
#L2k vs Safe ASO statistical testing

data_to_model2 <- expInputRBP[,pData(expInputRBP)$Treatment!='Toxic'] # or some equivalent filtering to remove the samples which are Vehicle treatment

#assign input and RBPs as factors in the material column of the data_to_model dataset.
material2 <- factor(data_to_model2$material, levels=c('Input', 'RBP'))

safe <- factor(data_to_model2$Treatment, levels=c('Safe', 'Vehicle'))


design2 <- model.matrix(~material2*safe)

# And then usual lines to perform linear modeling with limma...
```

Build the run_limma function
```{r}

run_limma2 <- function(data_to_model2,
                      design2,
                      coefficient){

  # run limma
  lmfit <- data_to_model2 %>%
    filterNA() %>%
    lmFit(design) %>%
    eBayes(trend=TRUE)

  # plot model vs mean variance
  plotSA(lmfit, main="Final model: mean-variance trend")

  # obtain results table
  lmResults2 <- topTable(lmfit, coef=coefficient, number=Inf)



  # add max_diff column if required
  if(is.integer(coefficient) & length(coefficient)>1){
    logfc_col='max_diff'
    message('Multiple numerical coffficients. Adding a "max_diff" column')
    lmResults2['max_diff'] <- lmResults2 %>%
      select(all_of(gsub(':', '.', colnames(coefficients(lmfit))[coefficient]))) %>%
      apply(MARGIN=1, function(x) x[which(abs(x)==max(abs(x)))])
  } else logfc_col='logFC'

  message('At 1% FDR')
  print(table(ifelse(lmResults$adj.P.Val<0.01, 'Significant', 'Not significant'),
              ifelse(lmResults[[logfc_col]]>0, 'Increase', 'Decrease')))

  message('At 20% FDR')
  print(table(ifelse(lmResults$adj.P.Val<0.2, 'Significant', 'Not significant'),
              ifelse(lmResults[[logfc_col]]>0, 'Increase', 'Decrease')))

  return(lmResults2)
}
```

build the volcano_plot function

```{r}
volcano_plot2 <- function(lmResultsObj2,
                         adj.P.Val_threshold=0.05,
                         logFC_threshold=2,
                         xcol='logFC',
                         xlabel){
  lmResultsObj2 %>%
    mutate(sig=adj.P.Val<adj.P.Val_threshold,
           large_effect=abs(!!sym(xcol))>logFC_threshold) %>%
    mutate(label=ifelse(sig & large_effect, gsub('(.*GN=| PE=.*)', '', Master.Protein.Descriptions), NA)) %>%
    ggplot(aes(!!sym(xcol), -log10(P.Value), colour=sig)) +
    geom_point(aes(size=sig)) +
    theme_camprot(base_size=15, border=FALSE) +
    scale_colour_manual(values=c('grey', get_cat_palette(1)), guide=FALSE) +
    scale_size_manual(values=c(0.5, 2), guide=FALSE) +
    xlab(xlabel) +
    ylab('-log10(p-value)') +
    theme(strip.background=element_blank()) +
    ggrepel::geom_label_repel(aes(label=label), show.legend=FALSE)
}
```

```{r}

lmResults2 <- run_limma2(data_to_model2, design2, coefficient='material2RBP:safeVehicle')

volcano2 <- volcano_plot(lmResults2,
                          adj.P.Val_threshold=0.1,
                          logFC_threshold=0.9,
                          xlabel='Effect of Safe ASO on RNA binding (log2)')

#define function build_results_list
build_results_list2 <- function(data_to_model2, design2, results2, volcano2){
  list('data'=data_to_model2,
       'design'=design2,
       'results2'=results2,
       'volcano'=volcano2)
}

lm_results2 <- build_results_list2(data_to_model2,
                                   design2,
                                   lmResults2,
                                   volcano2)
print(lm_results2$volcano2)

```

```{r}
fit1_2 <- lmFit(data_to_model2, design2)
#fit2 <- contrasts.fit(fit1,contrasts = contrast)
fit3_2 <- eBayes(fit1_2, trend = TRUE)


fit3_2$count = tibble::column_to_rownames(
   input_min_psm_count,'Master.Protein.Accessions')[common_proteins,'min_psm_count']

fit4_2 = spectraCounteBayes(fit3_2)
VarianceBoxplot(fit4_2,n=30,main="TMT15plex dataset",xlab="PSM count")


efit_deqms2 <- suppressWarnings(spectraCounteBayes(fit4_2))


# Diagnostic plots
VarianceBoxplot(efit_deqms2, n = 30, xlab = "PSMs")
```
```{r}

deqms_results2 <- outputResult(efit_deqms2, coef_col=4)


table(deqms_results2$sca.adj.pval<0.01)
#> 
#> FALSE  
#> 896


deqms_results2 %>%
  ggplot(aes(x = logFC, y = -log10(sca.P.Value), colour = sca.adj.pval < 0.01)) +
  geom_point(size=0.5) +
  theme_camprot(border=FALSE, base_size=15) +
  scale_colour_manual(values = c('grey', get_cat_palette(2)[2]), name = '2x vs 1x Sig.') +
  labs(x = 'Toxic vs Non-Toxic (Log2)', y = '-log10(p-value)') 


```
